---
title: 'Chapter 2: Inference using t-Distributions'
author: "Loren"
date: "08/17/2014"
output:
       knitrBootstrap::bootstrap_document:
        title: "Chapter 2: Inference using t-Distributions"
        theme: default
        highlight: xcode
        theme.chooser: TRUE
        highlight.chooser: TRUE
---

```{r bootstrap.show.code=FALSE, bootstrap.show.message=FALSE}
library(knitr)
opts_chunk$set(bootstrap.show.code=FALSE, bootstrap.thumbnail=FALSE, bootstrap.show.message=FALSE)
library(ggplot2)
theme_set(new = theme_minimal())
library(plyr)
library(reshape2)
# upper-case column names annoy me
sleuth.read <- function(x) {
   d <- read.csv(x)
   names(d) <- tolower(names(d))
   d
}
```

# Chapter 2: Inference using t-Distributions

## Bumpus sparrows

```{r}
case0201 <- sleuth.read('ascii/case0201.csv') # read the data
```

`r nrow(case0201)` live, stressed sparrows were collected after a storm.
Some survived and some died soon after.
(Don't ask me how they were collected).
Here is the number of birds in each condition:
```{r}
table(case0201$status)
```

The length of each bird's humerus was measured.
```{r fig.width=6, fig.height=2}
qplot(status, humerus, data=case0201, geom="boxplot") + coord_flip()
```

The two groups have about the same spread. Here is a t-test for
difference between group means, assuming equal variances:

```{r}
(bt <- t.test(humerus ~ status, data=case0201, var.equal=T))
```

There is no convincing evidence of a difference (p.val = `r bt$p.value`),
and removing the outliers would not increase the significance either.

## Schizophrenia in Identical Twins

```{r}
case0202 <- read.csv('ascii/case0202.csv')
```

The researchers found `r nrow(case0202)` pairs of identical twins, in
which one twin was affected by schizophrenia and the other was not.

The volume of a certain brain part (?) was measured in each twin.

In almost all pairs, the unaffected twin had higher volume:
```{r}
qplot(UNAFFECT - AFFECTED, data=case0202, binwidth=.125)
```

Here is a paired t-test:
```{r}
(test <- with(case0202, t.test(UNAFFECT, AFFECTED, paired=T)))
```
There is a significant difference (p-val = `r test$p.value`) in brain volume,
with an average difference of `r test$estimate`.  The 95% confidence interval
for the difference is `r test$conf.int[1]` to `r test$conf.int[2]`.

# Conceptual Exercises

1. If the dead birds were measured while they were dead, the experimenter could have been biased.
2. Because it's not what we're interested in?
3. "Shape of a histogram of the sample" -- false.  The histogram will be shaped like the population.
4. True

# Computational Exercises

## 12. Marijuana during pregnancy
```{r}
diff <- 280 # users - nonusers (birth weight)
se.diff <- 46.66
deg.free <- 1095
t.stat <- (diff - 0) / se.diff
```
a) 95% confidence interval for the difference between means:
```{r}
half.interval <- qt(.975,deg.free) * se.diff
c(-half.interval, half.interval) + diff
```
b) 90% confidence interval for the difference between means:
```{r}
half.interval <- qt(.95,deg.free) * se.diff
c(-half.interval, half.interval) + diff
```
c) 2-sided p-value for null hypothesis:
```{r}
2 * pt(-t.stat, deg.free)
```

## 13. Fish Oil and Blood Pressure
```{r}
ex0213 <- read.csv('ascii/ex0112.csv')
```
a) Group averages of BP:
```{r}
(aves <- with(ex0213, by(BP, DIET, FUN=mean)))
```
b) pooled estimate of standard deviation:
```{r}
(sizes <- table(ex0213$DIET))                # the group sizes
(sds <- with(ex0213, by(BP, DIET, FUN=sd)))  # the group standard deviations
(sp <- sqrt(sum((sizes - 1) * sds^2) / (sum(sizes -1)))) # the pooled standard deviation
```
c) SE(mean1 - mean2):
```{r}
(se <- sp * sqrt(sum(1/sizes)))
```
d) d.f. associated with sp is `r (my.df <- sum(sizes-1))`. The 97.5-th percentile of
the t-dist with this degree of freedom is `r qt(.975,my.df)`.
e) 

# Data Problems

## 21. Bumpus sparrow weight

```{r}
ex0221 <- sleuth.read('ascii/ex0221.csv')
```
We have the weights of `r nrow(ex0221)` sparrows, and whether they survived or not:
```{r}
head(ex0221)
```
Group sizes, boxplot, and t-test:
```{r}
summary(ex0221$status)
qplot(status, weight, data=ex0221, geom='boxplot') + geom_point()
t.test(weight ~status, data=ex0221)
```
T-test without "outlier":
```{r}
heaviest <- which.max(ex0221$weight)
t.test(weight ~status, data=ex0221[-heaviest,])
```
Whether the difference is "significant" (at the 0.05 level) depends on that
one data point, so it would be a very weak conclusion.

## 23. Speed limits and traffic fatalities
```{r}
ex0223 <- read.csv('ascii/ex0223.csv')
```
For each state, we have whether it raised its speed limit (over 55) in 1996,
and the percentage change in traffic fatalities from 1995 to 1996.
```{r}
head(ex0223)
```
```{r}
(table(ex0223$INCREASE))
qplot(FATALITIESCHANGE, data=ex0223, binwidth=10) + facet_grid(facets=INCREASE ~ .)
```

So, we have most of the states increasing their speed limit, and
the change in fatalities in seems to be higher among states that increased their
speed limit.  Yes, we have a significant t-test:
```{r}
(tt <- t.test(FATALITIESCHANGE ~ INCREASE, data=ex0223))
```
The confidence interval is (`r tt$conf.int`).
