---
title: 'Chapter 4: Alternatives to the t-Tools'
author: "Loren"
date: "09/06/2014"
output:
   knitrBootstrap::bootstrap_document:
        title: "Chapter 4: Alternatives to the t-Tools"
        theme: default
        highlight: xcode
        theme.chooser: TRUE
        highlight.chooser: TRUE
---

```{r bootstrap.show.code=FALSE, bootstrap.show.message=FALSE}
library(knitr)
opts_chunk$set(bootstrap.show.code=FALSE, bootstrap.thumbnail=FALSE, bootstrap.show.message=FALSE)
library(ggplot2)
theme_set(new = theme_minimal())
#use color brewer as default discrete colors
scale_colour_discrete <- function(...) scale_color_brewer(palette="Set1", ...)
scale_fill_discrete <- function(...) scale_fill_brewer(palette="Set1", ...)
library(plyr)
# upper-case column names annoy me
sleuth.read <- function(x) {
   d <- read.csv(x)
   names(d) <- tolower(names(d))
   d
}
```

# Conceptual Exercises


# Computational Exercises

## 14. O-Ring Study

In warm temperatures, there were 17 launches with 0 incidents, 2 launches with one incident, etc.

```{r}
ex0414 <- sleuth.read('ascii/case0401.csv')
# library(aplpack)
# with(ex0414, stem.leaf.backback(incidents[launch == 'COOL'], incidents[launch == 'WARM']))
with(ex0414, by(incidents, launch, FUN=function(x) { table(factor(x)) }))
# qplot(factor(incidents), data=ex0414, facets = launch ~ .)
```

```{r}
ttest <- t.test(incidents ~ launch, data=ex0414, var.equal=T)
```

A t-test on the data gives a p-value of `r ttest$p.val`, whereas the exact p-value
(from a permutation test) is 0.00988.

## 15. Artificial data: permutation test on differences of group means

```{r}
score <- c(1,5,4,8,9)
mean.diff <- mean(score[3:5]) - mean(score[1:2])
mean.diffs <- sort(aaply(combn(1:5,2), 2, function(x) { mean(score[-x]) - mean(score[x]) }))
names(mean.diffs) <- NULL
```

Group 1: 1, 5

Group 2: 4, 8, 9

Mean(group2) - Mean(group1) = `r mean.diff`.

The distribution of all such mean differences is [`r mean.diffs`].

`r (n <- sum(mean.diffs >= mean.diff))` of these are larger than or equal to `r mean.diff`,
for a one-sided p-value of `r n / length(mean.diffs)`.

## 16. More artificial data: permutation distribution of difference in sample averages

Group 1: 5, 7, 12

Group 2: 4, 6

```{r}
ex0416 <- c(5,7,12,4,6)
mean.diff <- mean(ex0416[1:3]) - mean(ex0416[4:5]) # group 1 minus group 2
dist <- sort(aaply(combn(1:5, 3), 2, function(x) { mean(ex0416[x]) - mean(ex0416[-x]) }))
names(dist) <- NULL
```

Mean(group1) - Mean(group2) = `r mean.diff`.

The distribution of such mean
differences, from all regroupings, is [`r dist`].

`r (num <- sum(abs(dist) >= mean.diff))` of these mean differences are equal to or more extreme
than `r mean.diff`, for a 2-sided p-value of `r num / length(dist)`.

## 18. A rank-sum test

The data:

Ranks 3,5, and 6 are in the treatment group.  1, 2, and 4 are in the control group.
The sum of ranks in the treatment group is `r sum(3,5,6)`.

Here is the distribution of possible rank sums for the treatment group, based on all permutations
of the ranks:

```{r fig.width=4, fig.height=2}
sum.trt <- sum(3,5,6)
rank.sums <- aaply(combn(1:6, 3), 2, sum)
qplot(rank.sums, data=data.frame(rank.sums)) +
   scale_x_continuous(breaks=6:15)
num.bigger <- sum(rank.sums >= sum.trt)
```

The graph shows that there are only `r num.bigger` rank sums bigger than or equal to `r sum.trt`.

## 19. Bumpus with rank-sum

Plot of the data:

```{r fig.width=7, fig.height=1.5}
ex0419 <- sleuth.read('ascii/case0201.csv')
ggplot(ex0419, aes(status, humerus)) + geom_boxplot() +
   theme(axis.title.y=element_blank()) + ylab("Humerus length (in.)") + coord_flip()
```

R's default rank-sum test uses a continuity correction.  It gives a warning that an exact p-value cannot be computed when there are ties:

```{r warning=T}
(wt <- wilcox.test(humerus ~ status, data=ex0419))
```

The reported 2-sided p-value, `r wt$p.value`, is much higher than
`r t.test(humerus ~ status, data=ex0419, var.equal=T)$p.value`, the value from a t-test,
but about the same as the p-value from a t-test in which the smallest observation has been
removed:
`r t.test(humerus ~ status, data=ex0419[-which.min(ex0419$humerus),])$p.value`.
TODO: interpret this.

## 20. Trauma and Metabolic Expenditure: rank-sum test "by hand"

```{r}
group <- c(rep('nontrauma', 8), rep('trauma', 7))
expenditure <- c(20.1, 22.9, 18.8, 20.9, 20.9, 22.7, 21.4, 20,
                           38.5, 25.8, 22, 23, 37.6, 30, 24.5)
ex0420 <- data.frame(expenditure, group)
ex0420 <- ex0420[order(ex0420$expenditure),]
```

<u>Metabolic expenditures (kcal/kg/day)</u>

Trauma patients: `r with(ex0420, sort(expenditure[group=='trauma']))`

Non-trauma patients: `r with(ex0420, sort(expenditure[group=='nontrauma']))`

```{r fig.width=7, fig.height=1.5}
ggplot(ex0420, aes(group, expenditure)) +
   geom_dotplot(binaxis='y', stackdir='center', dotsize=1.5, binwidth=.1) +
   theme(axis.title.y=element_blank()) +
   ylab("Metabolic expenditure (kcal/kg/day)") +
   coord_flip()
```

```{r}
ex0420$ranks <- rank(ex0420$expenditure)
```

```{r fig.width=7, fig.height=1.5}
ggplot(ex0420, aes(group, ranks)) +
   geom_dotplot(binaxis='y',stackdir='center', dotsize=.3) +
   theme(axis.title.y=element_blank()) + scale_y_discrete(breaks=1:15) +
   ylab("Rank of metabolic expenditure") +
   coord_flip()   
```

Rank sum for trauma group: `r (rank.sum.trauma <- with(ex0420, sum(ranks[group == 'trauma'])))`

Mean of all ranks: `r (mean.ranks <- mean(ex0420$ranks))`

Sd of all ranks: `r (sd.ranks <- sd(ex0420$ranks))`

Group sizes: `r (sizes <- table(ex0420$group))`

Theoretical mean of trauma rank-sum: `r (mean.T <- mean.ranks * sizes['trauma'])`

Theoretical Sd of trauma rank-sum: `r (sd.T <- sd.ranks * sqrt( prod(sizes) / sum(sizes) ))`

Z-statistic: `r (z <- (rank.sum.trauma - mean.T) / sd.T)`

**One-sided p-value**: `r (p <- pnorm(z, lower.tail=F))`

## 21. Trauma data: Checking exercise 20 using R's wilcox.test

```{r}
(wt <- wilcox.test(expenditure ~ group, data=ex0420, alternative = 'less'))
```

**Check: p-value** (with continuity correction) = `r wt$p.value`

## 22. Trauma data: 95% confidence interval for the difference of population medians

The set-up.  A function that tests whether a value is included in the interval:

```{r}
trauma <- with(ex0420, expenditure[group == 'trauma'])
nontrauma <- with(ex0420, expenditure[group == 'nontrauma'])

included <- function(delta) {   
   # 2-sided p-value, see page 93:
   wilcox.test(trauma, nontrauma + delta)$p.value > .05
}
included <- Vectorize(included) # Can do it for multiple test-deltas at once
```

Median(trauma) - Median(nontrauma) = `r median(trauma) - median(nontrauma) # 4.9`.

After some trial and error to find the endpoints:

```{r}
included(c(1.9, 1.91, 16.7, 16.71))
```

So the 95% confidence interval is approximately (1.91, 16.7).

## 23. Motivation and Creativity: randomization test compared to t-test

```{r fig.width=7, fig.height=1.5}
ex0423 <- sleuth.read('ascii/case0101.csv')
qplot(treatment, score, data=ex0423, geom='boxplot') +
   theme(axis.title.y=element_blank()) + coord_flip()
```

Group sizes:

```{r}
ex.size <- sum(ex0423$treatment == 'EXTRINSIC')
table(ex0423$treatment)
```


A randomization test:

```{r}
n <- 10000

mean.diffs <- numeric(n)
for (i in 1:n) {
   ex.ind <- sample(47, 23)
   mean.diffs[i] <- with(ex0423, mean(score[-ex.ind]) - mean(score[ex.ind]))
}
```

The observed difference in means:

```{r}
(in.minus.ex <- with(ex0423, mean(score[treatment=='INTRINSIC']) - mean(score[treatment=='EXTRINSIC'])))
```

Plot of the simulated differences in means:

```{r fig.width=6, fig.height=3}
hist(mean.diffs)
abline(v=c(in.minus.ex, -in.minus.ex), col='red')
```

Out of `r n` resamplings, only `r (blah <- sum(abs(mean.diffs) >= in.minus.ex))` gave
rise to a difference of means as extreme as the observed one, giving a 2-sided p-value
of `r blah / n`.

For comparison, the p-value of a t-test is
`r t.test(score ~ treatment, data=ex0423, var.equal=T)$p.value`.

## 24. Creativity and Motivation: 95% confidence interval using rank-sum procedure

```{r}
intrinsic <- with(ex0423, score[treatment == 'INTRINSIC'])
extrinsic <- with(ex0423, score[treatment == 'EXTRINSIC'])

included <- function(delta) {   
   # 2-sided p-value, see page 93:
   wilcox.test(intrinsic, extrinsic + delta)$p.value > .05
}
included <- Vectorize(included) # Can do it for multiple test-deltas at once
```

Median(intrinsic) - Median(extrinsic) = `r median(intrinsic) - median(extrinsic)`.

After some trial and error:

```{r}
included(c(1, 1.0001, 6.6, 6.601))
```

So the confidence interval is about (1, 6.6).  (That's for intrinsic minus extrinsic).

For comparison, a t-test gives a confidence interval of
`r t.test(score ~ treatment, data=ex0423, var.equal=T)$conf.int`.  (That's for
extrinsic minus intrinsic).

## 25. Guinea Pig Lifetimes: Welch t-test

```{r}
ex0425 <- sleuth.read('ascii/ex0211.csv')
wt <- t.test(lifetime ~ group, data=ex0425)
```

A Welch-test gives 2-sided p-value `r wt$p.value` and 95% confidence interval
(`r wt$conf.int`) for (bacilli - control).  However, the distribution is unusual.
The control group seems to be bimodal and the bacilli distribution is shaped
very differently.  The additive treatment model assumes that the treatment makes the same
difference (?), on each individual to which it is applied, hence that the
two distributions are shaped the same.  That doesn't seem to be the case here.

```{r fig.width=5, fig.height=3}
ggplot(ex0425, aes(lifetime)) +
   geom_histogram(binwidth=100, aes(y=..density..)) +
   # geom_density() + # is this just obnoxious?
   facet_grid(facets = group ~ .)
```

## 26. Schizophrenia - log-transforms

```{r}
ex0426 <- sleuth.read('ascii/case0202.csv')
```

The un-transformed differences in hippocampus volume:

```{r fig.width=5, fig.height=2}
ggplot(ex0426, aes(unaffect - affected)) + geom_histogram(binwidth=.05)
```

Are the data skewed? Perhaps, but only perhaps. Can you tell which QQ-plot
is the data?  (It's the top-center one).  The other plots are generated from
a normal distribution.

```{r fig.width=3, fig.height=2}
random.qq <- function() {
   x <- rnorm(15)
   qqnorm(x, main = FALSE, xaxt='n', yaxt='n', ann=FALSE)
}
par(mfrow = c(2,3))
par(mar=c(0,0,0,0))

random.qq()
qqnorm(ex0426$unaffect - ex0426$affected, main = FALSE, xaxt='n', yaxt='n', ann=FALSE)
random.qq()
random.qq()
random.qq()
random.qq()
```

The problem asks to log the volumes.

```{r}
ex0426$diff.of.logs <- with(ex0426, log(unaffect) - log(affected))
```

```{r}
t1 <- with(ex0426, t.test(unaffect - affected))
t2 <- with(ex0426, t.test(diff.of.logs))
```

For reference, a paired t-test on the untransfored data gives p-val = `r t1$p.value` and 95%
confidence interval (`r t1$conf.int`).  The mean difference in volume is estimated
to be `r t1$estimate`.

A paired t-test on the difference of the logged brain volumes gives p-val = `r t2$p.value`
and 95% confidence interval of (`r t2$conf.int`).  The mean difference in logged volume
is estimated to be `r t2$estimate`.  That means that the median of the ratios of the volumes
is estimated to be `r exp(t2$estimate)`, and the median of the ratios has a 95% confidence
interval of (`r exp(t2$conf.int)`).

## 27. TODO
