---
title: 'Chapter 5: Comparisons Among Several Samples'
author: "Loren"
date: "08/26/2014"
output:
   html_document:
      toc: true
      theme: spacelab
      highlight: pygments
---
```{r echo=F}
library(knitr)
opts_chunk$set(warning=FALSE, prompt=FALSE, message=FALSE)
library(ggplot2)
theme_set(new = theme_minimal())
library(plyr)
```

```{r echo=F}
# upper-case column names annoy me
sleuth.read <- function(x) {
   d <- read.csv(x)
   names(d) <- tolower(names(d))
   d
}

## Mice Diet and Longevity
```{r}
c0501 <- sleuth.read('ascii/case0501.csv')
```


## The Spock Trial

```{r echo=F}
c0502 <- sleuth.read('ascii/case0502.csv')
```

```{r}
require(ggplot2)
qplot(judge, percent, data=c0502, geom='boxplot')
table(c0502$judge)
```

Some models:

```{r}
full <- aov(percent ~ judge, c0502)
midway <- aov(percent ~ judge == "SPOCK'S", c0502)
reduced <- aov(percent ~ 1, c0502)
(test1 <- anova(full, reduced))
(test2 <- anova(midway, reduced)) # Significant at 1.03e-6
t.test(percent ~ (judge == "SPOCK'S"), c0502) # same p-value
```

A permutation test showing what the F-test means:

```{r}
n <- 1000
f.stats <- numeric(n)
my.fun <- function(x) { sum((x - mean(x))^2) }
sst <- my.fun(c0502$percent)           # total sum of squared resids
dft <- nrow(c0502) - 1                 # total degrees of freedom
dfb <- length(levels(c0502$judge)) - 1 # df between groups
dfw <- dft - dfb                       # df within groups
for (i in 1:n) {
   ssw <- sum(with(c0502, tapply(sample(percent), judge, my.fun)))
   msw <- ssw / dfw # mean squared resid, within groups
   msb <- (sst - ssw) / dfb
   f.stats[i] <- msb / msw
}
hist(f.stats, freq=F)
x <- seq(from=0,to=6,by=0.01)
y <- df(x,6,39)
lines(x,y)
```

Let me look more closely at the between/within sum of squares...

```{r}
my.fun <- function(x) { sum((x - mean(x))^2) }
sst <- my.fun(c0502$percent)
df.total <- nrow(c0502) - 1
df.between <- length(levels(c0502$judge)) - 1
df.within <- df.total - df.between
ssw <- sum(with(c0502, tapply(percent, judge, my.fun))) # SS within groups

grand.mean <- mean(c0502$percent)
group.means <- with(c0502, tapply(percent, judge, mean))
group.sizes <- with(c0502, tapply(percent, judge, length))
group.resids <- group.means - grand.mean

# the same:
sst - ssw
weighted.mean((group.means - grand.mean)^2, group.sizes) * nrow(c0502)
```

## Diagnostics using Residuals

```{r echo=F}
c0501.full <- aov(lifetime ~ diet, data=c0501)
plot(c0501.full) # not sure how to get separate plots
df <- data.frame(estimated.means=c0501.full$fitted.values,
                 residuals=c0501.full$residuals)
```

Plotting means vs residuals:

```{r echo=F}
ggplot(df, aes(estimated.means, residuals)) +
   geom_point() + xlab('Estimated Mean Lifetime (months)') +
   ylab('Residual Lifetime (months)') +
   geom_hline(yintercept=0)
```

# Conceptual Exercises

1. 
2. 
3. 

# Computational Exercises

## 13. Spock Trial continued

```{r}
ex0513 <- read.csv('ascii/case0502.csv')
names(ex0513) <- tolower(names(ex0513))
```

a) What is the average percentage of women from all venires?

```{r}
(ave.women <- mean(ex0513$percent))
```

b) For how many of the 9 Spock judge's venires is the percentage of women less than this grand average?  All 9:

```{r}
with(ex0513, sum(percent < ave.women & judge == "SPOCK'S"))
```

c) For how many of the venires of Spock's judge is the percentage of women less than the average percentage for Spock's judge?

```{r}
sum(with(ex0513[ex0513$judge == "SPOCK'S",], percent < mean(percent)))
```

## 14. Spock - "by hand" t-test of Spock's judge vs judge A

a) What is the pooled standard deviation?

```{r}
spocks.percent <- 14.62
spocks.sd <- 5.039
spocks.n <- 9
a.n <- 5
a.percent <- 34.12
a.sd <- 11.942
(sd.pooled <- sqrt(((spocks.n-1)*spocks.sd^2 + (a.n-1)*a.sd^2)/
                      (spocks.n + a.n - 2)))
```

b) t-test

```{r}
df.test <- spocks.n + a.n - 2
(se <- sd.pooled * sqrt(1/spocks.n + 1/a.n)) # standard error of (a - spock)
(t <- (0 - (a.percent - spocks.percent)) / se)
2 * pt(t, df=df.test)
```

c) Checking against a computer t-test:

```{r}
ind <- with(ex0513, judge == "SPOCK'S" | judge == "A")
t.test(percent ~ judge, data=ex0513[ind,], var.equal=T)
```

## 15. More Spock

a) sample variance

```{r}
(sv <- var(ex0513$percent))
```

b) RSS

```{r}
45 * sv
with(ex0513, sum((percent - mean(percent))^2) ) # a check
```

c) TODO

## 16. Spock computer F-test

```{r}
full <- aov(percent ~ judge, data=ex0513)
reduced <- aov(percent ~ 1, data=ex0513)
(model.comparison <- anova(full, reduced))
```

## 17. An F-test by hand - TODO: do the table in markdown style

<table border="1">
<tr>
<th>Source</th><th>df</th><th>Sum of squares</th><th>Mean square</th><th>F-stat</th><th>p-val</th>
</tr>
<tr>
<td>Between Groups</td> <td>7</td>  <td>35819</td> <td>5117</td>  <td>3.5</td>   <td>0.01424</td>
</tr>
<tr>
<td>Within groups</td>  <td>24</td> <td>35088</td> <td>1462</td>
</tr>
<tr>
<td>Total</td>          <td>31</td> <td>70907</td>
</tr>
</table>

## 18. Fatty Acid

```{r}
ex0518 <- read.csv('ascii/ex0518.csv')
names(ex0518) <- tolower(names(ex0518))
```

A model assuming six independent groups, and plots of residual vs. day.

```{r}
six <- aov(protein ~ treatmnt, ex0518)
plot(six$residuals ~ ex0518$day, main="All treatments", ylab="residuals", xlab="day")
with(ex0518, plot(six$residuals[ treatmnt != 'CONTROL'] ~ day[treatmnt != 'CONTROL'],
                  main="The non-control groups", ylab="residuals", xlab="day"))
with(ex0518, plot(six$residuals[ treatmnt == 'CONTROL'] ~ day[treatmnt == 'CONTROL'],
                  main="Just the control groups.", ylab="residuals", xlab="day"))
```

The residuals in the "just control" group are calculated in reference to the mean of the whole
control group.  So the plot shows that the observations jump around **together** from day to day.
So the day seems to have an affect on the protein measurement.

The ten groups do not have equal means [shorthand for "we reject the null hypothesis that
the ten groups have equal means"]:

```{r}
ten <- aov(protein ~ trt.daygroup, ex0518)
anova(ten)
```

The means of the control group change from day to day:
[Why are the two p-values different?]

```{r}
anova(ten, six)
anova(aov(protein ~ day, data=subset(ex0518, treatmnt=='CONTROL')))
```

## 19. Cavity Size and Use - TODO - don't understand this one

```{r}
species <- c('mouse', 'pinyon mouse', "Bewick's wren", 'mountain bluebird',
             'ash-throated flycatcher', 'plain titmouse', 'northern flicker',
             'western screech owl', 'american kestrel')
n <- c(127,44,24,41,18,16,11,7,6)
mn <- c(7.347,7.368,7.418,7.487,7.563,7.568,8.214,8.272,8.297)
sample.sd <- c(.4979,.4235,.3955,.3183,.3111,.4649,.2963,.3242,.5842)
ex0519 <- data.frame(species, n, mn, sample.sd)
```

a) Compute the pooled estimate of variance.

```{r}
(s2p <- with(ex0519, weighted.mean(sample.sd^2, n - 1))) # see pg. 120-121
```

b) F-test

```{r}
extra.df <- (9-1)
ssw <- with(ex0519, sample.sd^2 %*% (n-1))   # SS within groups
sst <- 0.4962^2 * (294-1)                # SS total (0.4962 is their given SD)
ssb <- sst - ssw                       # SS between groups
f <- (ssb / extra.df) / s2p
pf(f, df1 = extra.df, df2 = (294-1) - (9-1), lower.tail = F)
```

c) Verify another way of calculating ssb.

d) Sneaky way of checking whether my answers above were correct: making fake
data and having R do the F-test.

```{r}
sneaky.sample <- function(n, desired.sample.mean, desired.sample.sd) {
   x <- rnorm(n)
   x <- (x - mean(x)) / sd(x)
   desired.sample.sd * x + desired.sample.mean
}
require(plyr)
set.seed(5432)
fake.data <- ddply(ex0519, .(species), transform, sam = sneaky.sample(n, mn, sample.sd))
anova(aov(sam ~ species, data=fake.data))
```

## Tufte plot of Spock data

```{r fig.width=5, fig.height=2.5}
c0502 <- sleuth.read('ascii/case0502.csv')
ggb <- ggplot_build(ggplot(c0502, aes(judge, percent)) + geom_boxplot())
ggplot(ggb$data[[1]]) + geom_segment(aes(x=x, y=ymin, xend=x, yend=lower)) +
   geom_segment(aes(x=x, y=upper, xend=x, yend=ymax)) +
   geom_point(aes(x=x, y=middle)) + xlab("Judge") + ylab("Percent women") +
   ggtitle("Tufte 'box' plot of Spock data\n(outliers not included!)") +
   coord_flip()
```


TODO

# Data Problems

## 23. Was T-Rex warm-blooded?

```{r}
ex0523 <- read.csv('ascii/ex0523.csv')
names(ex0523) <- tolower(names(ex0523))
```

Sure look different to me:

```{r}
ggplot(ex0523, aes(bonegrp, oxygen)) + geom_boxplot() + theme_minimal()
```

The spreads look pretty different though.

```{r}
anova(aov(oxygen ~ bonegrp, data=ex0523))
```

## 24. Zinc and vegetarians (observational)

```{r}
ex0524 <- read.csv('ascii/ex0524.csv')
names(ex0524) <- tolower(names(ex0524))
```

Doesn't look like much evidence:

```{r echo=F}
ggplot(ex0524, aes(group, zinc)) + geom_boxplot()
```

