---
title: 'Chapter 5: Data Problems'
author: "Loren"
date: "08/26/2014"
output:
   knitrBootstrap::bootstrap_document:
        title: "Chapter 5: Data Problems"
        theme: default
        highlight: xcode
        theme.chooser: TRUE
        highlight.chooser: TRUE
---


```{r bootstrap.show.code=FALSE, bootstrap.show.message=FALSE}
library(knitr)
opts_chunk$set(bootstrap.show.code=FALSE, bootstrap.thumbnail=FALSE, bootstrap.show.message=FALSE)
library(ggplot2)
theme_set(new = theme_minimal())
#use color brewer as default discrete colors
scale_colour_discrete <- function(...) scale_color_brewer(palette="Set1", ...)
scale_fill_discrete <- function(...) scale_fill_brewer(palette="Set1", ...)
library(plyr)
# upper-case column names annoy me
sleuth.read <- function(x) {
   d <- read.csv(x)
   names(d) <- tolower(names(d))
   d
}
```


# Data Problems

## 23. Was T-Rex warm-blooded?

```{r}
ex0523 <- read.csv('ascii/ex0523.csv')
names(ex0523) <- tolower(names(ex0523))
```

Sure look different to me:

```{r}
ggplot(ex0523, aes(bonegrp, oxygen)) + geom_boxplot() + theme_minimal()
```

The spreads look pretty different though.

```{r}
anova(aov(oxygen ~ bonegrp, data=ex0523))
```

## 24. Zinc and vegetarians (observational)

```{r}
ex0524 <- read.csv('ascii/ex0524.csv')
names(ex0524) <- tolower(names(ex0524))
```

Doesn't look like much evidence:

```{r}
ggplot(ex0524, aes(group, zinc)) + geom_boxplot()
```


The residuals in the "just control" group are calculated in reference to the mean of the whole
control group.  So the plot shows that the observations jump around **together** from day to day.
So the day seems to have an affect on the protein measurement.

The ten groups do not have equal means [shorthand for "we reject the null hypothesis that
the ten groups have equal means"]:

```{r}
ex0518 <- sleuth.read('ascii/ex0518.csv')
six <- aov(protein ~ treatmnt, ex0518)
ten <- aov(protein ~ trt.daygroup, ex0518)
anova(ten)
```

The means of the control group change from day to day:
(Why are the two p-values different?)

```{r}
anova(ten, six)
anova(aov(protein ~ day, data=subset(ex0518, treatmnt=='CONTROL')))
```
