---
title: 'Chapter 5: Computational Exercises'
author: "Loren"
date: "08/26/2014"
output: html_document
---

knitrBootstrap::bootstrap_document:
        title: "Chapter 5: Computational Exercises"
        theme: default
        highlight: xcode
        theme.chooser: TRUE
        highlight.chooser: TRUE

i hate things that don't work.  ```{r bootstrap.show.code=FALSE, bootstrap.show.message=FALSE}
```{r}
library(knitr)
# opts_chunk$set(bootstrap.show.code=FALSE, bootstrap.thumbnail=FALSE, bootstrap.show.message=FALSE)
library(ggplot2)
theme_set(new = theme_minimal())
library(plyr)
# upper-case column names annoy me
sleuth.read <- function(x) {
   d <- read.csv(x)
   names(d) <- tolower(names(d))
   d
}
```

# Chapter 5: Computational Exercises

## 13. Spock Trial continued

```{r}
ex0513 <- sleuth.read('ascii/case0502.csv')
```

a) What is the average percentage of women from all venires?

```{r}
(ave.women <- mean(ex0513$percent))
```

b) For how many of the 9 Spock judge's venires is the percentage of women less than this grand average?  All 9:

```{r}
with(ex0513, sum(percent < ave.women & judge == "SPOCK'S"))
```

c) For how many of the venires of Spock's judge is the percentage of women less than the average percentage for Spock's judge?

```{r}
sum(with(ex0513[ex0513$judge == "SPOCK'S",], percent < mean(percent)))
```

## 14. Spock - "by hand" t-test of Spock's judge vs judge A

a) What is the pooled standard deviation?

```{r}
spocks.percent <- 14.62
spocks.sd <- 5.039
spocks.n <- 9
a.n <- 5
a.percent <- 34.12
a.sd <- 11.942
(sd.pooled <- sqrt(((spocks.n-1)*spocks.sd^2 + (a.n-1)*a.sd^2)/
                      (spocks.n + a.n - 2)))
```

b) t-test

```{r}
df.test <- spocks.n + a.n - 2
(se <- sd.pooled * sqrt(1/spocks.n + 1/a.n)) # standard error of (a - spock)
(t <- (0 - (a.percent - spocks.percent)) / se)
2 * pt(t, df=df.test)
```

c) Checking against a computer t-test:

```{r}
ind <- with(ex0513, judge == "SPOCK'S" | judge == "A")
t.test(percent ~ judge, data=ex0513[ind,], var.equal=T)
```


## 15. More Spock

a) sample variance

```{r}
(sv <- var(ex0513$percent))
```

b) RSS

```{r}
45 * sv
with(ex0513, sum((percent - mean(percent))^2) ) # a check
```

c) TODO

## 16. Spock computer F-test

```{r}
full <- aov(percent ~ judge, data=ex0513)
reduced <- aov(percent ~ 1, data=ex0513)
(model.comparison <- anova(full, reduced))
```



## 17. An F-test by hand

TODO

## 18. Fatty Acid

```{r}
ex0518 <- sleuth.read('ascii/ex0518.csv')
```

A model assuming six independent groups, and plots of residual vs. day.

```{r}
six <- aov(protein ~ treatmnt, ex0518)
plot(six$residuals ~ ex0518$day, main="All treatments", ylab="residuals", xlab="day")
with(ex0518, plot(six$residuals[ treatmnt != 'CONTROL'] ~ day[treatmnt != 'CONTROL'],
                  main="The non-control groups", ylab="residuals", xlab="day"))
with(ex0518, plot(six$residuals[ treatmnt == 'CONTROL'] ~ day[treatmnt == 'CONTROL'],
                  main="Just the control groups.", ylab="residuals", xlab="day"))
```


The residuals in the "just control" group are calculated in reference to the mean of the whole
control group.  So the plot shows that the observations jump around **together** from day to day.
So the day seems to have an affect on the protein measurement.

The ten groups do not have equal means (shorthand for "we reject the null hypothesis that
the ten groups have equal means"):

```{r}
ten <- aov(protein ~ trt.daygroup, ex0518)
anova(ten)
```


The means of the control group change from day to day:
(Why are the two p-values different?)

```{r}
anova(ten, six)
anova(aov(protein ~ day, data=subset(ex0518, treatmnt=='CONTROL')))
```


## 19. Cavity Size and Use - TODO - don't understand this one

```{r}
species <- c('mouse', 'pinyon mouse', "Bewick's wren", 'mountain bluebird',
             'ash-throated flycatcher', 'plain titmouse', 'northern flicker',
             'western screech owl', 'american kestrel')
n <- c(127,44,24,41,18,16,11,7,6)
mn <- c(7.347,7.368,7.418,7.487,7.563,7.568,8.214,8.272,8.297)
sample.sd <- c(.4979,.4235,.3955,.3183,.3111,.4649,.2963,.3242,.5842)
ex0519 <- data.frame(species, n, mn, sample.sd)
```

a) Compute the pooled estimate of variance.

```{r}
(s2p <- with(ex0519, weighted.mean(sample.sd^2, n - 1))) # see pg. 120-121
```


b) F-test

```{r}
extra.df <- (9-1)
ssw <- with(ex0519, sample.sd^2 %*% (n-1))   # SS within groups
sst <- 0.4962^2 * (294-1)                # SS total (0.4962 is their given SD)
ssb <- sst - ssw                       # SS between groups
f <- (ssb / extra.df) / s2p
pf(f, df1 = extra.df, df2 = (294-1) - (9-1), lower.tail = F)
```




c) Verify another way of calculating ssb.

d) Sneaky way of checking whether my answers above were correct: making fake
data and having R do the F-test.

```{r}
sneaky.sample <- function(n, desired.sample.mean, desired.sample.sd) {
   x <- rnorm(n)
   x <- (x - mean(x)) / sd(x)
   desired.sample.sd * x + desired.sample.mean
}
require(plyr)
set.seed(5432)
fake.data <- ddply(ex0519, .(species), transform, sam = sneaky.sample(n, mn, sample.sd))
anova(aov(sam ~ species, data=fake.data))
```

## Tufte plot of Spock data

```{r fig.width=5, fig.height=2.5}
c0502 <- sleuth.read('ascii/case0502.csv')
ggb <- ggplot_build(ggplot(c0502, aes(judge, percent)) + geom_boxplot())
ggplot(ggb$data[[1]]) + geom_segment(aes(x=x, y=ymin, xend=x, yend=lower)) +
   geom_segment(aes(x=x, y=upper, xend=x, yend=ymax)) +
   geom_point(aes(x=x, y=middle)) + xlab("Judge") + ylab("Percent women") +
   ggtitle("Tufte 'box' plot of Spock data\n(outliers not included!)") +
   coord_flip()
```


TODO
