---
title: "Chapter 6: Linear Combinations and Multiple Comparisons of Means"
author: "Loren Serfass"
date: "09/07/2014"
output:
   knitrBootstrap::bootstrap_document:
        title: "Chapter 6: Linear Combinations and Multiple Comparisons of Means"
        theme: default
        highlight: xcode
        theme.chooser: TRUE
        highlight.chooser: TRUE
---

```{r bootstrap.show.code=FALSE, bootstrap.show.message=FALSE}
library(knitr)
opts_chunk$set(bootstrap.show.code=FALSE, bootstrap.thumbnail=FALSE, bootstrap.show.message=FALSE)
library(ggplot2)
theme_set(new = theme_minimal())
library(plyr)
# upper-case column names annoy me
sleuth.read <- function(x) {
   d <- read.csv(x)
   names(d) <- tolower(names(d))
   d
}
```

# Computational Exercises

## 12. Handicap Study: linear contrast of disabilities of mobility vs disability of communication

```{r echo=F}
ex0612 <- sleuth.read('ascii/case0601.csv')
```

My function to do a contrast with estimated gamma of 0.
*x* is the numerical variable, *index* is the grouping
variable.  *group1* and *group2* are each vectors of level-numbers.

```{r}
sd.pooled <- function(x, index) {
   sd.devs <- tapply(x, index, sd)
   sizes <- tapply(x, index, length)
   sqrt(weighted.mean(sd.devs^2, sizes - 1))
}
   
linear.contrast <- function(x, index, group1, group2) {
   means <- tapply(x, index, mean)
   sizes <- tapply(x, index, length)
   sdp <- sd.pooled(x, index) # see pg. 121
   d.f <- sum(sizes - 1)
   
   J <- length(group1)
   K <- length(group2)
   # g <- sum(means[group1]) / J - sum(means[group2]) / K # one way to do it, see pg. 156
   # or this way:
   coeffs <- numeric(5)
   coeffs[group1] <- 1 / J
   coeffs[group2] <- -1 / K
   g <- coeffs %*% means # see pg. 154
   
   # finding a 95% confidence interval
   se.g <- sdp * sqrt(coeffs^2 %*% (1/sizes)) # see pg. 154
   t.ratio <- g / se.g
   p.value <- 2 * pt(abs(t.ratio), df=d.f, lower.tail = F)
   t <- qt(.975, d.f) # t-value for upper end of 95% confidence interval
   conf.int <- g + se.g * c(-t, t)
   
   l <- list()
   l$p.value <- p.value
   l$conf.int <- conf.int
   l$coeffs <- coeffs
   l$g <- g
   l$se.g <- se.g
   l
}
```

Using the function to re-create the contrast between wheelchair/crutches and amputee/hearing
(in the book on pg 155):

```{r}
levels(ex0612$handicap) # hearing/crutches = c(2,5), amputee/hearing = c(1,3)
cont <- with(ex0612, linear.contrast(score, handicap, c(2,5), c(1,3)))
```

The p-value is `r cont$p.value` and the confidence interval is (`r cont$conf.int`).

Next, a contrast of mobility handicaps vs hearing impairment:

```{r}
cont <- with(ex0612, linear.contrast(score, handicap, c(1,2,5), 3))
```

The p-value is `r cont$p.value` and the confidence interval is (`r cont$conf.int`).

## 13. Handicap study - Bonferroni method

The question asks for simultaneous confidence intervals for these differences:

* Crutches - Amputee (2 - 1)
* Amputee - Wheelchair (1 - 5)
* Crutches - Wheelchair (2 - 5)

```{r}
ind.conf.level <- 1 - .05/3
attach(ex0612)
sp <- sd.pooled(score, handicap)
sizes <- tapply(score, handicap, length)
se <- sp * sqrt(sum(1/sizes))
detach(ex0612)
```

