---
title: 'Chapter 7: Simple Linear Regression: A Model for the Mean'
author: "Loren Serfass"
date: "09/12/2014"
output:
       knitrBootstrap::bootstrap_document:
        title: "Chapter 7: Simple Linear Regression: A Model for the Mean"
        theme: default
        highlight: xcode
        theme.chooser: TRUE
        highlight.chooser: TRUE
---

# Chapter 7: Simple Linear Regression: A Model for the Mean

```{r bootstrap.show.code=FALSE, bootstrap.show.message=FALSE}
library(knitr)
opts_chunk$set(bootstrap.show.code=FALSE, bootstrap.thumbnail=FALSE, bootstrap.show.message=FALSE)
library(ggplot2)
theme_set(new = theme_minimal())
library(plyr)
sleuth.read <- function(x) {
   d <- read.csv(x)
   names(d) <- tolower(names(d))
   d
}
```

## Computational Exercises

### 12. Var and SD in linear regression

```{r}
residuals <- c(-3.3,2.1,-4,-1.5,5.1,1.6)
(sd.dev <- sum(residuals^2) / (length(residuals) - 2)) # Standard deviation.
```

The sample standard deviation is $\hat{\sigma} = `r sd.dev`$, so the variance is
$\hat{\sigma}^2 = `r sd.dev^2`$.  The degrees of freedom are `r length(residuals) - 2`.

### 13. Big bang: confidence interval for intercept

```{r}
c0701 <- sleuth.read('ascii/case0701.csv')
m <- lm(distance ~ velocity, data=c0701)
sm <- summary(m)
se.intercept <- sm$coefficients[1,2]
half.width <- qt(.975, m$df.residual) * se.intercept
sm$coefficients[1,1] + c(-half.width, half.width)
```

### 14. Planet distances linear model

```{r fig.width=5, fig.height=3.5}
ex0714 <- sleuth.read('ascii/ex0116.csv')
# with(ex0714, plot(order, distance))
ggplot(ex0714, aes(order, log(distance))) + geom_point() +
   geom_smooth(method='lm', se=F) +
   geom_text(aes(label=as.character(planet)),hjust=-.4,size=3.5) +
   scale_x_continuous(limits=c(0,12), breaks=0:12)
m <- lm(log(distance) ~ order, ex0714)
# log(ex0714$distance) == m$fitted.values + m$residuals
```

These are the estimates of intercept and slope:

```{r}
m$coefficients
```

Calculating the fitted values and residuals (these are also already
in the m object as `m$fitted.values` and `m$residuals`):

```{r}
my.fitted <- m$coefficients[1] + ex0714$order * m$coefficients[2]
my.resid <- log(ex0714$distance) - my.fitted
```

Calculating estimate of sigma:

```{r}
sum(my.resid^2) / 8
```

### 15. Planet distance and Bode's law

```{r}
sm <- summary(m)
coef <- sm$coefficients[2,]
t <- (coef['Estimate'] - log(2)) / coef['Std. Error']
names(t) <- NULL
(p <- 2 * pt(abs(t), df=sm$df[2], lower.tail = F)) # TODO: is this right??
```

The p-value for the hypothesis that $\beta_{1} = log(2)$ is p.val = `r p`, so we reject
that hypothesis.


$SD(T) = S_{R} \sqrt{\dfrac{n_{1}n_{2}}{(n_{1}+n_{2})}}$

Sample correlation coefficient:

$r_{XY} = \dfrac{\sum\limits_{i=1}^{n}(   X_{i} - \overline{X})(Y_{i} - \overline{Y})/(n-1)}{s_{X}s_{Y}}$

### 23.

```{r echo=F}
ex0723 <- sleuth.read('ascii/ex0723.csv')

```

### 24. Crab claw size and force

```{r echo=F, fig.height=3, fig.width=7}
ex0724 <- sleuth.read('ascii/ex0724.csv')
ggplot(ex0724, aes(log(height), log(force), color=species)) + geom_point() +
   geom_smooth(method='lm', se=F)
```

```{r}
models <- dlply(ex0724, 'species',
                function(d) { lm(log(force) ~ log(height), data=d) } )
```

### Digression: contouring the cost function of a linear regression

```{r}
n <- 50
x <- rnorm(n, 10, 4)
y <- 3 + 0.5 * x + rnorm(n, 0, 2)
d <- data.frame(x,y)
ggplot(d, aes(x,y)) + geom_point() + geom_smooth(method='lm')
sm <- summary(lm(y ~ x, d))
ci.seq <- function(est, se, df) {
   hw <- se * qt(.975,df)
   seq(from = est - hw, to = est + hw, length.out = 50)
}
beta0 <- ci.seq(sm$coefficients[1,1], sm$coefficients[1,2], sm$df[2])
beta1 <- ci.seq(sm$coefficients[2,1], sm$coefficients[2,2], sm$df[2])
cost <- Vectorize(function(b0, b1) { (1/(2*n)) * sum( (b0 + b1* d$x - d$y)^2 ) })
betas <- expand.grid(beta0=beta0, beta1=beta1)
betas$cost <- cost(betas$beta0, betas$beta1)
ggplot(betas, aes(beta0,beta1)) + geom_contour(aes(z=cost),bins=50) +
   annotate(geom="point", x=sm$coefficients[1,1], y=sm$coefficients[2,1])
ggplot(betas, aes(beta0,beta1)) + geom_tile(aes(fill=cost)) +
   geom_contour(aes(z=cost),bins=20,color='red')
```

