---
title: 'Chapter 8: A Closer Look at Assumptions for Simple Linear Regression'
author: "Loren Serfass"
date: "02/12/2015"
output:
   knitrBootstrap::bootstrap_document:
        title: "Chapter 8: A Closer Look at Assumptions for Simple Linear Regression"
        theme: default
        highlight: xcode
        theme.chooser: TRUE
        highlight.chooser: TRUE
---

# Chapter 8: A Closer Look at Assumptions for Simple Linear Regression

```{r bootstrap.show.code=FALSE, bootstrap.show.message=FALSE}
library(knitr)
opts_chunk$set(bootstrap.show.code=FALSE, bootstrap.thumbnail=FALSE, bootstrap.show.message=FALSE)
library(ggplot2)
theme_set(new = theme_minimal())
#use color brewer as default discrete colors
scale_colour_discrete <- function(...) scale_color_brewer(palette="Set1", ...)
scale_fill_discrete <- function(...) scale_fill_brewer(palette="Set1", ...)
library(plyr)
sleuth.read <- function(x) {
   d <- read.csv(x)
   names(d) <- tolower(names(d))
   d
}
```

# This chapter's examples

## Island area and number of species

This dataset gives the area and number of animal species (reptile and amphibian)
of 7 islands in the West Indies. Here is the whole dataset:

```{r}
(case0801 <- sleuth.read('ascii/case0801.csv'))
```

Plotting the variables directly does not result in a straight line.

```{r fig.width=3, fig.height=3}
qplot(area, species, data=case0801)
```

The log-log plot is very close to a straight line though.

```{r fig.width=3, fig.height=3}
qplot(log(area), log(species), data=case0801) + geom_smooth(method='lm')
```

## Breakdown times for insulating fluid under different voltages

Here, logging the y scale results in a fairly linear relationship.

```{r}
case0802 <- sleuth.read('ascii/case0802.csv')
ybreaks <- array((1:9) %*% t(10^(0:3)))
ylabs <- character(36)
ylabs[seq(1,by=9,length=4)] <- 10^(0:3)
qplot(voltage, time, data=case0802) + scale_y_log10(breaks=ybreaks, labels=ylabs) +
   geom_smooth(method='lm') +
   scale_x_continuous(breaks=seq(26,38,by=2)) + ggtitle("Breakdown times") + ylab("Time (in minutes)")
```

# Chapter notes

Linear models assume: *linearity, constant variance, normality, and independence*.
If the true relationship between the variables is far enough from linear, then
predictions are *biased*, over- or under- estimating. ...
Violations of the *normality* assumption don't cause a lot of inaccuracy
in the coefficient estimates, but prediction intervals will be inaccurate.
If observations are *not independent*, cluster and serial effects should be
incorporated into models.

```{r}
par(mfrow=c(2,2))
x <- runif(200,0,200); plot(x, .5*x+10*rnorm(200), main="use linear regression")
x <- rexp(200,1); plot(x, .5*log(x)+.5*rnorm(200), main="transform X")
x <- runif(200,-2,2); plot(x, -x^2 + .5*rnorm(200), main="include X^2")
x <- runif(200,0,200); plot(x, exp(.01*x + .3*rnorm(200)), main="transform Y (log, reciprocal, sqrt)")
par(mfrow=c(1,1))
```

Plotting residuals vs fitted values to assess a model.  First, modeling the untransformed `time`.

```{r fig.width=10, fig.height=3.5}
bad.model <- lm(time ~ voltage, data=case0802)
par(mfrow=c(1,2))
plot(case0802$voltage, case0802$time, ylab="breakdown time", xlab="voltage"); abline(bad.model)
plot(bad.model$fitted.values, bad.model$residuals); abline(h=0)
```

Trying the square root of breakdown time:

```{r fig.width=10, fig.height=3.5}
sqrt.model <- lm(sqrt(time) ~ voltage, data=case0802)
par(mfrow=c(1,2))
plot(case0802$voltage, sqrt(case0802$time), ylab="sqrt(breakdown time)", xlab="voltage"); abline(sqrt.model)
plot(sqrt.model$fitted.values, sqrt.model$residuals); abline(h=0)
```

Trying the log of breakdown time.  The variance in this model seems fairly constant.

```{r fig.width=10, fig.height=3.5}
log.model <- lm(log(time) ~ voltage, data=case0802)
par(mfrow=c(1,2))
plot(case0802$voltage, log(case0802$time), ylab="log(breakdown time)", xlab="voltage"); abline(log.model)
plot(log.model$fitted.values, log.model$residuals); abline(h=0)
```

Plotting the three models on the same plot. On the log scale, the black and red lines
don't continue because they go below 0.

```{r fig.width=10, fig.height=4}
par(mfrow=c(1,2))
plot(case0802$voltage, case0802$time, ylab="breakdown time", xlab="voltage")
legend(x="topright", lwd=1, col=c('black', 'red', 'blue'), legend=c('untransformed','sqrt','log'), box.lty=0)
abline(bad.model)
lines(case0802$voltage, sqrt.model$fitted.values^2, col='red')
lines(case0802$voltage, exp(log.model$fitted.values), col='blue')
plot(case0802$voltage, case0802$time, ylab="breakdown time", xlab="voltage", log="y", main="log scale")
legend(x="topright", lwd=1, col=c('black', 'red', 'blue'), legend=c('untransformed','sqrt','log'), box.lty=0)
lines(case0802$voltage, bad.model$fitted.values, col='black')
lines(case0802$voltage, (sqrt.model$fitted.values^2)*ifelse(sqrt.model$fitted.values>0,1,-1), col='red')
lines(case0802$voltage, exp(log.model$fitted.values), col='blue')
```

## Interpretationg after log transformations

When only Y is logged: *use medians*. When X increases by 1, Y's median is multiplied.

When X is logged: a [doubling of] X is associated with adding a constant amount to Y.

When both X and Y are logged: a [doubling of X] is associated with multiplying Y's median by a constant factor.